### 2016-08-31 robust-software.tex
Prettier home page

Typically, scientific software is developed and used by a single person,
usually a graduate student or postdoc, and produces excellent results in their
hands\cite{prins2015}. 
But what happens
when someone else wants to use their software? Everyone with
a few years of experience feels a tremor of fear when told to use
a graduated student's code to analyze their data.
Often, that software will be undocumented and work in unexpected
ways (if it works at all).  It will often rely on nonexistent paths or resources,
be tuned for a single dataset,
or simply be an older version than was used in published papers.
The new user is then faced with two unpalatable options:
hack the existing code to make it work, or start over. Being unable to
replicate results is so common that one publication refers to it as "a rite of
passage"\cite{baker2016}.

The root cause of this problem is that most research software
isn't \emph{robust}. The difference between running and being
robust is the difference between ``works for me on my machine'' and
``works for other people on a cluster I've never used''. With a lack of
robustness also comes a lack of reproducibility and many duplicated efforts,
slowing the pace of research\cite{prabhu2011,lawlor2015}.
Bioinformatics is full of such efforts, catalogued by efforts such as
the ELIXIR tools and data registry\cite{ison2016}
and the Bioinformatics Links Directory \cite{brazas2012}. In the Bioinformatics
Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
to analyze transcript expression, and 182 pathway and interaction resources.

This problem is not unique to bioinformatics, or even to computing\cite{baker2016}. 
Best practices in software engineering specifically aim to increase software
robustness. However, most bioinformaticians come from related disciplines and
are trained on-the-job or otherwise informally\cite{prins2015,atwood2015}. Even
the existing programs and initiatives rarely have the time to cover engineering
topics in-depth, especially since the field is so broad and rapidly
developing\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
and knowledge are taught, developing robust software is not directly rewarded
in science and funding is difficult to come by\cite{prins2015}. Some proposed
solutions to this problem, in addition to restructuring educational programs,
include hiring dedicated software engineers \cite{lawlor2015,sanders2008},
partnering with private sector or grassroots organizations\cite{prins2015,ison2016},
or using specific technical tools like containerization or cloud
computing\cite{afgan2016,howe2012}. Each of these requires time and, in some
cases, institutional change.

You don't need to be a full-time programmer to write robust software. In fact,
some of the best, most reliable pieces of software in many scientific
communities are written by researchers \cite{prabhu2011,sanders2008}. 
The most widely-used software packages tend to adopt strong software
engineering approaches, have high standards of reproducibility, good testing
practices, and inter-institutional collaborations that grow into strong user
bases. In the bioinformatics community, Bioconductor and Galaxy follow this
path\cite{Gentleman2004,afgan2016}.

That said, not \emph{every} coding effort requires such rigor:
as the saying goes, not everything worth doing is worth doing right (or right away).
Code that is used once to answer a specific question related to a specific dataset
doesn't require comprehensive documentation or flexible configuration,
and the only sensible way to test it may well be to run it on the dataset in
question. Exploratory analysis is an iterative process that is developed
quick and revised often\cite{lawlor2015,sanders2008}. 
However, if a script is dusted off and run three or four
times for slightly different purposes,
is crucial to a publication or a lab,
or being passed on to someone else,
it may be time to make your software more robust.
In particular, robust software:

>>>>

Typically, scientific software is developed and used by a single person,
usually a graduate student or postdoc, and produces excellent results in their
hands~\cite{prins2015}. 
But what happens
when someone else wants to use their software? Everyone with
a few years of experience feels a tremor of fear when told to use
a graduated student's code to analyze their data.
Often, that software will be undocumented and work in unexpected
ways (if it works at all).  It will often rely on nonexistent paths or resources,
be tuned for a single dataset,
or simply be an older version than was used in published papers.
The new user is then faced with two unpalatable options:
hack the existing code to make it work, or start over. Being unable to
replicate results is so common that one publication refers to it as "a rite of
passage"~\cite{baker2016}.

The root cause of this problem is that most research software
isn't \emph{robust}. The difference between running and being
robust is the difference between ``works for me on my machine'' and
``works for other people on a cluster I've never used''. With a lack of
robustness also comes a lack of reproducibility and many duplicated efforts,
slowing the pace of research~\cite{prabhu2011,lawlor2015}.
Bioinformatics is full of such efforts, catalogued by efforts such as
the ELIXIR tools and data registry~\cite{ison2016}
and the Bioinformatics Links Directory~\cite{brazas2012}. In the Bioinformatics
Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
to analyze transcript expression, and 182 pathway and interaction resources.

This problem is not unique to bioinformatics, or even to computing~\cite{baker2016}. 
Best practices in software engineering specifically aim to increase software
robustness. However, most bioinformaticians come from related disciplines and
are trained on-the-job or otherwise informally~\cite{prins2015,atwood2015}. Even
the existing programs and initiatives rarely have the time to cover engineering
topics in-depth, especially since the field is so broad and rapidly
developing~\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
and knowledge are taught, developing robust software is not directly rewarded
in science and funding is difficult to come by~\cite{prins2015}. Some proposed
solutions to this problem, in addition to restructuring educational programs,
include hiring dedicated software engineers~\cite{lawlor2015,sanders2008},
partnering with private sector or grassroots organizations~\cite{prins2015,ison2016},
or using specific technical tools like containerization or cloud
computing~\cite{afgan2016,howe2012}. Each of these requires time and, in some
cases, institutional change.

You don't need to be a full-time programmer to write robust software. In fact,
some of the best, most reliable pieces of software in many scientific
communities are written by researchers~\cite{prabhu2011,sanders2008}. 
The most widely-used software packages tend to adopt strong software
engineering approaches, have high standards of reproducibility, good testing
practices, and inter-institutional collaborations that grow into strong user
bases. In the bioinformatics community, Bioconductor and Galaxy follow this
path~\cite{gentleman2004,afgan2016}.

That said, not \emph{every} coding effort requires such rigor:
as the saying goes, not everything worth doing is worth doing right (or right away).
Code that is used once to answer a specific question related to a specific dataset
doesn't require comprehensive documentation or flexible configuration,
and the only sensible way to test it may well be to run it on the dataset in
question. Exploratory analysis is an iterative process that is developed
quick and revised often~\cite{lawlor2015,sanders2008}. 
However, if a script is dusted off and run three or four
times for slightly different purposes,
is crucial to a publication or a lab,
or being passed on to someone else,
it may be time to make your software more robust.
In particular, robust software:
